---
title: "Fundermentals of R"
subtitle: '5.0 R Operations Assignement'
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# SetUp

Define the following variables: -

- a with the value of 10
- b with the value of 5
- c with the value of TRUE
- d with the value of "hello"

# Question 1

Using Arithmetic Operators,calculate the following:-

    1. a + b
    2. a - b
    3. a * b
    4. a / b
    5. a ^ 2
    
```{r}
# your code
```


# Question 2

Using comparison operators, evaluate the following expressions and store the result in a variable

    1. `a` is equal to `b`
    2. `a` is not equal to `b`
    3. `a` is greater than `b`
    4. `a` is less than `b`
    5. `a` is greater than or equal to `b`
    6. `a` is less than or equal to `b`
    
```{r}
# your code
```    
    
# Question 3

Using logical operators, evaluate the following expressions and store the result in a variable:

    1. `c` AND `c`
    2. `c` OR `c`
    3.  NOT `c`
    4. `a` is greater than `b` AND `c` is TRUE
    5. `d` is equal to "hello" OR `c` is FALSE
    
```{r}
# your code
```

# Question 4

Using concatenation operator, combine the following vectors

    x <- c("apple", "banana", "orange")
    y <- c("grape", "pineapple", "watermelon")
    
```{r}
# your code
```

# Question 5

Write a program that prompts the user to enter two numbers and performs the following operations:-

    1. Adds the two numbers
    2. Subtracts the second number from the first number
    3. Multiplies the two numbers
    4. Divides the first number by the second number
    
```{r}
# your code
```

# Question 6

Using the seq() function, create a vector of integers from 1 to 10. Using arithmetic operators, calculate the sum of the vector.

```{r}
# your code
```

# Question 7

Create a variable age with a value of 30. Using comparison operators, write an if-else statement that prints "You are an adult" if age is greater than or equal to 18, and "You are not yet an adult" if age is less than 18.

```{r}
# your code
```

# Question 8

Using logical operators, write a program that prompts the user to enter a number and checks if it is a multiple of both 3 and 5. Print "The number is a multiple of both 3 and 5" if it is, and "The number is not a multiple of both 3 and 5" if it is not

```{r}
# your code
```

# Question 9

Using the concatenation operator, create a vector fruits with the values "apple", "banana", and "orange". Create another vector prices with the values 0.50, 0.30, and 0.75. Using the paste() function, combine the two vectors into a new vector fruit_prices, where each element is a combination of the fruit name and its corresponding price (e.g. "apple - 0.50"). Print the resulting vector.

```{r}
# your code
```

# Question 10

Using the rep() function, create a vector with 10 repetitions of the string "hello". Using the `toUpper()` function, convert the vector to all uppercase letters. Using the concatenation operator, add an exclamation mark to the end of each element in the vector. Print the resulting vector.

```{r}
# your code
```

# NOTE
Your submission should include a well-documented R code explaining each step of your code.