---
title: "Fundermentals of R"
subtitle: "Advanced R - 4.0 Machine Learning - Logistic Regression"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# loading libraries
library(tidyverse)
library(mice) # missing data imputation
library(viridis) # fill plots with colors
library(caret) # data splitting
library(sjPlot) # print model summary as a plot
```

## Introduction

Logistic regression is a widely used algorithm for binary classification problems in machine learning. In this lesson, we will explore how to implement machine learning using logistic regression in R. We will cover the entire process, from loading and cleaning the data to model training, evaluation, and making predictions.

## The Data

The sinking of the Titanic is one of the most infamous shipwrecks in history.

On April 15, 1912, during her maiden voyage, the widely considered "unsinkable" RMS Titanic sank after colliding with an iceberg. Unfortunately, there weren't enough lifeboats for everyone onboard, resulting in the death of 1502 out of 2224 passengers and crew.

While there was some element of luck involved in surviving, it seems some groups of people were more likely to survive than others.

In this challenge, we will build a predictive model that answers the question: "what sorts of people were more likely to survive?" using passenger data (ie name, age, gender, socio-economic class, etc). We will use the titanic dataset in this lesson, you can learn more from the kaggle website.

## Data Loading

The dataset consists of the train and test datasets. We will use the training dataset to train and test our model and the best model will be used to make final predictions using the test set.

The data dictionary:

| Variable | Definition                                | Key                                            |
|---------------|---------------------------|------------------------------|
| Survival | Survival                                  | 0 = No, 1 = Yes                                |
| pclass   | Ticket class                              | 1 = 1st, 2 = 2nd, 3 = 3rd                      |
| Sex      | Sex                                       | male, female                                   |
| Age      | Age in years                              |                                                |
| sibsp    | \# of siblings/spouses aboard the Titanic |                                                |
| parch    | \# of parents/children aboard the Titanic |                                                |
| ticket   | Ticket number                             |                                                |
| fare     | Passenger fare                            |                                                |
| cabin    | Cabin number                              |                                                |
| embarked | Port of Embarkation                       | C = Cherbourg, Q = Queenstown, S = Southampton |

: Data Dictionary

```{r data-loading}
# loading the train data
train = read.csv("data/train.csv")

# subset columns you need
train_updated = train %>%
  select(-c(PassengerId, Name, SibSp, Parch, Ticket, Cabin, Embarked))

# data summary
summary(train_updated)

# na count
colSums(is.na(train_updated))
```

## Data Cleaning

### Missing Data

```{r handle-na}
# define the imputation model
imp_model <- mice(train_updated, method = "pmm", m = 5)

# get complete data
imputed_data <- complete(imp_model)

colSums(is.na(imputed_data))
```

### Data Types

The data types of all variables are wrongly labeled except for Age and Fare. We will convert the data types of Survived, PClass and Sex into factors.

```{r data-types}
# converting data types to factors
imputed_data = imputed_data %>%
  mutate(
    Survived = as.factor(Survived),
    Pclass = as.factor(Pclass),
    Sex = as.factor(Sex)
  )


# data summary
summary(imputed_data)
```

## Exploratory Data Analysis

### Survival rate by passenger class

```{r surv-by-pclass}
imputed_data %>%
  mutate(
    Survived = ifelse(Survived == 0, "No", "Yes"),
    Pclass = ifelse(Pclass == 1, "Upper", 
                    ifelse(Pclass == 2, "Middle", "Lower"))
  ) %>%
  ggplot(aes(x = Pclass, fill = Survived)) +
  geom_bar(position = "dodge") +
  scale_fill_viridis(discrete = T) +
  labs(title = "Survival rate by passenger class",
       subtitle = "Data Source: Kaggle Titanic dataset",
       x = "Passenger class",
       y = "Survival rate",
       fill = "Survived") +
  theme_bw()

```

This plot shows the survival rate for passengers in each class. As you can see, passengers in first class had the highest survival rate, followed by passengers in second class. Passengers in third class had the lowest survival rate.

### Age Distribution of Survivors

```{r survivors-age-dist}
imputed_data %>%
  mutate(
    Survived = ifelse(Survived == 0, "No", "Yes")
  ) %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Age)) +
  geom_histogram(binwidth = 10) +
  labs(title = "Age Distribution of Survivors",
       x = "Age",
       y = "Number of Survivors") +
  theme_bw()
```

This plot shows the age distribution of survivors. As you can see, there were more survivors in the younger age groups. This is likely because children and young adults were more likely to be given priority for lifeboats.

### Fare Paid by Passenger Class

```{r fare-by-pclass}
imputed_data %>%
  mutate(
    Pclass = ifelse(Pclass == 1, "Upper", 
                    ifelse(Pclass == 2, "Middle", "Lower"))
  ) %>%
  ggplot(aes(x = Pclass, y = Fare)) +
  geom_boxplot() +
  labs(title = "Fare Paid by Passenger Class",
       x = "Passenger Class",
       y = "Fare") +
  theme_bw()
```

This plot shows the distribution of fare paid by passenger class. As you can see, passengers in first class paid the highest fares, followed by passengers in second class. Passengers in third class paid the lowest fares.

### Survival Rate by Sex

```{r surv-by-sex}
imputed_data %>%
  mutate(
    Survived = ifelse(Survived == 0, "No", "Yes")
  ) %>%
  ggplot(aes(x = Sex, fill = Survived)) +
  geom_bar(position = "dodge") +
  scale_fill_viridis(discrete = T) +
  labs(title = "Survival Rate by Sex",
       x = "Sex",
       y = "Survival Rate",
       fill = "Survived") +
  theme_bw()
```

This plot shows the survival rate for passengers by sex. As you can see, female passengers had a higher survival rate than male passengers. This is likely because women and children were more likely to be given priority for lifeboats.

### Survival Rate by Age and Sex

```{r surv-age-sex}
imputed_data %>%
  mutate(
    Survived = ifelse(Survived == 0, "No", "Yes")
  ) %>%
  ggplot(aes(x = Age, fill = Survived)) +
  geom_density(alpha = 0.2) +
  facet_wrap(~Sex) +
  labs(title = "Survival Rate by Age and Sex",
       x = "Age",
       y = "Density",
       fill = "Survived")

```

This plot shows the survival rate for passengers by age and sex. As you can see, female passengers had a higher survival rate than male passengers at all ages. However, the survival rate for female passengers was particularly high in the younger age groups.

## Modelling

### Splitting the Data

We will split the data into: -

      1. Training Data (80%).
      2. Validation Data (20%).
      
```{r data-splitting}
# splitting index
index = createDataPartition(imputed_data$Survived, p=0.8, list = FALSE)

# training data
training = imputed_data[index,]

# validation data
validation = imputed_data[-index,]
```

### Fitting Base model

```{r}
# the model
model = glm(Survived~., data = training, family = binomial(link="logit"))

# model summary
tab_model(model)
```

### Predictions and Evaluation

```{r pred}
# predictions
pred = predict(model, newdata = validation, type = "response")

# picking a threshold of 0.6
pred = ifelse(pred >= 0.6, 1, 0)

# confusion matrix
mt = table(validation$Survived,pred)

confusionMatrix(mt)
```

The base model achieved an accuracy of `r confusionMatrix(mt)$overall[['Accuracy']]`.

### Visualize results

To plot the results of a logistic regression model, you can create various types of visualizations depending on the specific requirements and objectives. Here are a few common ways to visualize the results of a logistic regression model in R:

      **ROC Curve**
      
    The Receiver Operating Characteristic (ROC) curve is a widely used graphical          representation to evaluate the performance of a binary classifier. It shows the       trade-off between the true positive rate (sensitivity) and the false positive         rate (1 - specificity) at various classification thresholds.
    
```{r roc-curve}
# ROC curve
library(pROC)
roc_curve <- roc(validation$Survived, pred)
plot(roc_curve, main = "ROC Curve", col = "blue")
```


    **Coefficient Plot**
      
     A coefficient plot can be used to visualize the estimated coefficients of the        logistic regression model, providing insights into the importance and direction of     each variable's impact on the outcome.`
    
```{r coef-plot}
coef_data <- data.frame(variable = names(coef(model)), coefficient = coef(model))

ggplot(coef_data, aes(x = variable, y = coefficient)) + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_bar(stat = "identity", fill = "blue") + 
  coord_flip() +
  labs(title = "Logistic Regression Coefficient Plot") +
  theme_bw()
```

CONGRATULATIONS OF FINISHING YOUR LESSON, PLEASE PROCEED TO THE ASSIGNMENT.
