---
title: "Fundermentals of R"
subtitle: "Intermediate R - 2.0 Statistical Analysis Assignment"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set-Up

Create a data frame named `df` with the variables `Student`, `Age`, `gender`, `maths`, `pass` and fill it with 20 records.

The variables contain the following data types:

      Student => string
      Age => number
      gender => factor
      maths => number
      pass => boolean
      
```{r set-up}
# your code
```

## Question 1

Print-out the summary of `df.`

```{r q1}
# your code
```

## Question 2

Who is the youngest and oldest student?

```{r q2}
# your code
```

## Question 3

What percentage of the students are male and female?

```{r q3}
# your code
```

## Question 4

What percentage of students passed the exams?

```{r q4}
# your code
```

## Question 5

Write a function without using the built function and use it to calculate the mean maths performance?

```{r q5}
# your code
```

## Question 6

Plot the distribution of the students performance for maths.

```{r q6}
# your code
```

## Question 7

generate the frequency distribution table for both `gender` and `pass`.

```{r q7}
# your code
```

## Question 8

What is the performance range for maths?

```{r q8}
# your code
```

## Question 9

Add a new column, `Young`, which contains a Boolean. The Boolean should be `TRUE` if the student is younger than 20 years.

```{r q9}
# your code
```

## Question 10

Write a brief overview of the data summary.

```{r q10}
# your code
```

NICE ONE! PLEASE PROCEED TO THE NEXT LESSON.
