---
title: "Fundermentals of R"
subtitle: "Intermediate R - 2.0 Statistical Analysis"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

R provides a powerful set of functions and packages for statistical analysis. In this lesson, we will cover various statistical measures and analysis techniques in R, including calculating minimum, maximum, standard deviation, quartiles, mean, mode, median, summary statistics, range, frequency distribution, and interquartile range.

## Descriptive Statistics

### Minimun Value

```{r min}
# create the vector we will work with
vec = c(1,2,3,4,-20,100,230)

# lowest value
min(vec)
```

### Maximum Value

```{r max}
# highest value
max(vec)
```

### Standard Deviation

```{r sd}
# Standard Deviation
sd(vec)
```

### Quartiles

```{r qt}
# quatiles
quantile(vec)
```

### Average

```{r average}
# average
mean(vec, na.rm = TRUE)
```

### Median

```{r median}
# median
median(vec, na.rm = TRUE)
```

### Range

```{r range}
# range
range(vec, na.rm = TRUE)
```

### Interquartile Range

```{r iqr}
# Interquartile range
IQR(vec, na.rm = TRUE)
```

## Frequency Distribution

A frequency distribution displays the frequency (count) of each unique value in a data set.

```{r fq}
# freaquency
table(vec)
```


