---
title: "Fundermentals of R"
subtitle: '3.0 Conditional Statements'
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Conditional statements are used to  handle decisions in R. They are used to perform certain actions if a certain criteria is met. In this session, we will cover the following :- 

    1. If statements.
    2. If else statements.
    3. Else if statements.
    4. Next statements.
    5. Break Statements.
    6. While loop.
    7. For loop.
    
# If statements

An if statement is only executed if the condition being tested evaluates to TRUE. The code chunks below will illustrate some of the use cases.

## Greater than

```{r}
# check if a number if greater than a given value and print it

# lets define the variables
num = 10

# the if statement
if(num > 10){
  cat(num)
}

# since the num is not greater than 10 (evaluates to FALSE), the code is not 
# executed
```

## Equal to

```{r}
# check if num is equal to
if(num == 10){
  cat(num)
}
```

## Not Equal to

```{r}
# check if num is not equal to
if(num != 10){
  cat(num)
}
```

## Greater than or Equal to

```{r}
# check if num is greater than or equal to
if(num >= 10){
  cat(num)
}
```

## Less than or Equal to

```{r}
# check if num is greater than or equal to
if(num <= 10){
  cat(num)
}
```

# If Else Statements

An if else statements is used to handle situations that are not handled by the if statement. For instance, in the if statement, if the num is not equal to 10, the code will not be executed, to solve this we will use the else statement to give the user feedback.

Let use create a new variable, name, and use it to illustrate if else statements.

```{r}
# defining name
name = "james bond"

# let check if the name jane doe is equal to name
if(name == "jane doe"){
  cat(name)
} else {
  cat("name is not jane doe")
}
```

Lets change the name from *jane doe* to *james bond*.

```{r}
if(name == "james bond"){
  cat(name)
} else {
  cat("name is not jane doe")
}
```

As seen above, if the if statement does not catch the condition, it is forwarded to the else condition. Lets try the same but with numbers.

```{r}
# declare num2
num2 = 2

# test if the number is greater than 2
if(num2 > 2){
  cat("number is greater than 2")
} else {
  cat("number is less than 2")
}
```

