---
title: "Fundermentals of R"
subtitle: '3.0 Conditional Statements'
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Conditional statements are used to  handle decisions in R. They are used to perform certain actions if a certain criteria is met. In this session, we will cover the following :- 

    1. If statements.
    2. If else statements.
    3. Else if statements.
    4. Next statements.
    5. Break Statements.
    6. While loop.
    7. For loop.
    
# If statements

An if statement is only executed if the condition being tested evaluates to TRUE. The code chunks below will illustrate some of the use cases.

```{r}
# check if a number if greater than a given value and print it

# lets define the variables
num = 10

# the if statement
if(num > 10){
  num
}

# since the num is not greater than 10 (evaluates to FALSE), the code is not 
# executed
```


```{r}
# check if num is equal to
if(num == 10){
  num
}
```


```{r}
# check if num is not equal to
if(num != 10){
  num
}
```


```{r}
# check if num is greater than or equal to
if(num >= 10){
  num
}
```