---
title: "Fundermentals of R"
subtitle: '3.0 Conditional Statements'
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Conditional statements are used to  handle decisions in R. They are used to perform certain actions if a certain criteria is met. In this session, we will cover the following :- 

    1. If statements.
    2. If else statements.
    3. Else if statements.
    4. AND and OR.
    5. Next statements.
    6. Break Statements.
    7. While loop.
    8. For loop.
    
# If statements

An if statement is only executed if the condition being tested evaluates to TRUE. The code chunks below will illustrate some of the use cases.

## Greater than

```{r}
# check if a number if greater than a given value and print it

# lets define the variables
num = 10

# the if statement
if(num > 10){
  cat(num)
}

# since the num is not greater than 10 (evaluates to FALSE), the code is not 
# executed
```

## Equal to

```{r}
# check if num is equal to
if(num == 10){
  cat(num)
}
```

## Not Equal to

```{r}
# check if num is not equal to
if(num != 10){
  cat(num)
}
```

## Greater than or Equal to

```{r}
# check if num is greater than or equal to
if(num >= 10){
  cat(num)
}
```

## Less than or Equal to

```{r}
# check if num is greater than or equal to
if(num <= 10){
  cat(num)
}
```

# If Else Statements

An if else statements is used to handle situations that are not handled by the if statement. For instance, in the if statement, if the num is not equal to 10, the code will not be executed, to solve this we will use the else statement to give the user feedback.

Let use create a new variable, name, and use it to illustrate if else statements.

```{r}
# defining name
name = "james bond"

# let check if the name jane doe is equal to name
if(name == "jane doe"){
  cat(name)
} else {
  cat("name is not jane doe")
}
```

Lets change the name from *jane doe* to *james bond*.

```{r}
if(name == "james bond"){
  cat(name)
} else {
  cat("name is not jane doe")
}
```

As seen above, if the if statement does not catch the condition, it is forwarded to the else condition. Lets try the same but with numbers.

```{r}
# declare num2
num2 = 2

# test if the number is greater than 2
if(num2 > 2){
  cat("number is greater than 2")
} else {
  cat("number is less than 2")
}
```

# Else If Statements

Else if statements helps one to handle multiple conditions before executing the else condition.

The code chunk below illustrates how to make use of the else if statements, we will use the name variable.

```{r}
# check if the name is available and give access
name = "james bond"

if(name == "james bond"){
  cat("name exists in the data base")
} else if(name == "Clinton Bond"){
  cat("name clinton exists")
} else {
  cat("name does not exists")
}
```

In the code chunk above, condition 1 is met because the name is james bond.

In the code chunk below, we change the name to clinton Bond, observe the results.

```{r}
# check if the name is available and give access
name = "Clinton Bond"

if(name == "james bond"){
  cat("name exists in the data base")
} else if(name == "Clinton Bond"){
  cat("name clinton exists")
} else {
  cat("name does not exists")
}
```

# AND and OR statements.

The AND and OR statements are logical operators that are used to combine multiple conditional statements. They help one to check for multiple criteria before executing a give code.

When using AND, the code only executes if both conditions are true and when using OR, the code executes even when at-least one of the conditions in true.

The code chunks below demonstrate some of the use cases.

## AND operator

```{r}
# define the variables
a = 5
b = 7

# check if both are greater than
if(a > 4 & b > 5){
  cat("both a and b are greater than")
} else {
  cat("both a and b are less than")
}

```

## OR operator

```{r}
# check if any is greater than
if(a > 5 | b > 5){
  cat("atleast one greater than")
} else {
  cat("atleast one is less than")
}

```