---
title: "Fundermentals of R"
subtitle: '3.0 Conditional Statements Assignements'
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Set Up Q1

Create two variables,`x = 10 and y = 20`, and tackle the following question.

## Question 1

Create a new variable, z, and assign it to the result of y > x. Print the value of z using the  `cat()` function

```{r}
# your code goes here
```

Change the value of z to your full name and print the new value of z.

## Question 2

```{r}
# your code goes here
```

# Set Up Q2

Create a data frame, **Perf**, containing the performance of your class. The data frame is already created for you in the code chunk below, print it.

```{r}
# creating the perf data frame
perf = data.frame(
  Student = c("James", "Peter", "Andrew", "Mercy",  "Magdalene"),
  Math = c(89,79,50,40,98)
)
```

## Question 1

Create a new column, **Grade**, containing the grades of each student based on the following grading system.

    0 - 30 - D
    31 - 50 - C
    51 - 70 - B
    71 - 80 - B+
    81 - 100 - A
    
```{r}
# your code goes here
```

## Question 2

Create a new column, **Pass**, containing a Boolean value on whether a student has passed or not. *The pass mark is 50*.

```{r}
# your code goes here
```

# Set Up Q3

Create an empty vector, **vec**, of length 10. 

## Question 1

Print the contents of *vec*.

```{r}
# your code goes here
```

Create another vector, **vec2**, containing the data `1,2,3,4,5,6,7,8,9,10`. Using the appropriate conditional statements, fill up **vec** with twice the value of **vec2**.

## Question 2

Print the values of **vec**.

*Hint: Use the functions length and for loop*

```{r}
# your code goes here
```



