---
title: 'Beginner: Fundermentals Of R'
subtitle: '2.0 Data Structures'
author: "Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Data structure is the format in which data is organized, managed and stored for easy storage and access. In R, there are many formats in which data can be stored, some of them include: 

    1. Vectors.
    2. Lists.
    3. Matrices.
    4. Arrays.
    5. Data Frames.

In this lesson, we will discuss the above data structures and as a bonus, we will also cover **Factors**, **Strings**, **comments** and **variables**.

# Vectors

A vector is a data structure that is used to store data elements of the same data type. They are arrays with one-dimension.

## Creating Vectors

In R, the following methods can be used to create vectors: -

1. Using the c() function.
2. Using the seq() function.
3. Using the colon (:) for numerical values

### Using the c() function.

To create vectors using the concatenate function (c()), you will use the format `vec_name = c(sequency of numbers/strings)` as illustrated in the code chunkS below.

```{r}
# creating a numerical vectors
vec = c(1,2,3,4)
vec
```

`Hint: To run a code chunk, click on the green play button at the top right of the code chunk.`

```{r}
# creating a vector with character data types
vec2 = c("R","Python","C++","JavaScript")
vec2
```


```{r}
# creating a vector with logical data types
vec3 = c(TRUE,FALSE,FALSE,TRUE)
vec3
```

### Using the seq() function.

```{r}
# creating numerical vector with seq
vec4 = seq(0,5)
vec4

# creating numerical vector with seq with a difference of 2 between each values
vec5 = seq(0,5, by=2)
vec5

# creating a vector with a defined length
# trying changing the length.out and observe
vec6 = seq(0,5, length.out=10)
vec6
```

### Using the colon operator

```{r}
# creating a numerical vector with values from 1 to 10
vec7 = 1:10
vec7

# creating a numerical vector with values from -1 to 10
vec8 = -1:10
vec8
```


`TIP: WHEN YOU LOOK oN YOUR ENVIRONMENT TAB, YOU SHOULS BE ABLE TO SEE ALL THE VECTORS WE HAVE CREATED`

## Vector Operations.

There are various operations that can applied to vectors, they include **vector length**, **vector arithmetic**, **combining vector** and **vector indexing**

### Vector Length

To find out the number of items in a vector, we use the `length()` function. For instance, we can use this function to find out how many items in the vectors we have created above as shown below, run the code chunk below to get the length of the vectors.

```{r}
# length of vec, vec4 and vec8
paste0("Length of vec: ", length(vec))
paste0("Length of vec2: ", length(vec4))
paste0("Length of vec8: ", length(vec8))
```
### Vector Indexing

The items of a vector can be accessed by referring to its index. The index of a vector start from index 1 which extracts the first item. *Tip: you can use the `length` function to get the index of the last item*

Run the following code chunks to access various items from the vector.

```{r}
# access the first item of vec
vec[1]
```


```{r}
# access the last item of vec
vec[4]
```


```{r}
# access the first and last item of vec
vec7[c(1,length(vec7))]
```


```{r}
# range indexing
# extract 2nd to 3rd values
vec2[c(2:4)]
```

### Vector arithmetics

Vectors in R can be added can be added, subtracted, divided and multiplied provided they have the same length. The items of a vector can be multiplied, subtracted, divided and added by a certain value,k.

Run the code chunks below to see the vector arithmetics in action.

```{r}
# add 5 to vec
vec + 5
```


```{r}
# subtract 5 from vec
vec - 5
```


```{r}
# multiply 5 to vec
vec * 5
```


```{r}
# adding vec and vec4
vec6 + vec7

# subtract vec and vec4
vec6 - vec7

# multiply vec and vec4
vec6 * vec7

# divide vec and vec4
vec6 / vec7
```


### Sorting Vectors

The elements of a vector can be arranged in either ascending or descending order. The code chunks below show how sorting can be in vectors, Run the code chunks below.

```{r}
# creating anew vector
vec = c(20, 4, 100, 56)

vec

# sorting the vector vec in ascending order
sort(vec)

# sorting the vector vec in descending order
sort(vec, decreasing = TRUE)
```


# Bonus

## Comments

To write comments in R, you will use `#` followed by the code. The code chunk below illustrates how to write code. 

```{r}
# this is a comment, it will not be executed.
```


