---
title: 'Beginner: Fundermentals of R'
subtitle: '2.0 Data Structures Assignement'
author: "Prepared By Clinton Moshe Mwachia"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instrictions.

      1. Comment on what your code does.
      2. Make sure the code only prints the answer.

# Vectors

## PREPARATION.

Create two vectors, **animals**, containing the big five animals in KENYA and **marks** containing the math scores of five candidates. Use the code chunk below. 

```{r prep}

```

## Question 1

What is the length of the vector **animals**?

write your code below.

```{r}

```

## Question 2

What is the length of the vector **marks**?

write your code below.

```{r}

```

## Question 3

Extract the 1st, 3rd and 5th animals and print them to the console.

```{r}

```

## Question 4

What is the highest math score of the candidates?

*hint: sort the marks in descending order and extract the first item*

```{r}

```

# Lists

## Preparation.

create a list, **my_list** that contains the two vectors you created earlier.

```{r}

```

## Question 1

What is the length of the list you have created above?

```{r}

```
## Question 2

create a new vector, **vec** containing numbers 100 and 200 and append it to **my_list**.

What is the length of the new list?

```{r}

```

## Question 3

The new list created above does not have names.Proceed to the code chunk below and appropriately name the items. Print the new list.

```{r}

```

## Question 4

Extract the 1st, 3rd and 5th animals from the list above and print them to the console

```{r}

```

