---
title: 'Beginner: Fundermentals Of R'
subtitle: '2.0 Data Structures'
author: "Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Data structure is the format in which data is organized, managed and stored for easy storage and access. In R, there are many formats in which data can be stored, some of them include: 

    1. Vectors.
    2. Lists.
    3. Matrices.
    4. Arrays.
    5. Data Frames.

In this lesson, we will discuss the above data structures and as a bonus, we will also cover **Factors**, **Strings**, **comments** and **variables**.

# Vectors

A vector is a data structure that is used to store data elements of the same data type. They are arrays with one-dimension.

## Creating Vectors

In R, the following methods can be used to create vectors: -

1. Using the c() function.
2. Using the seq() function.
3. Using the colon (:) for numerical values

### Using the c() function.

To create vectors using the concatenate function (c()), you will use the format `vec_name = c(sequency of numbers/strings)` as illustrated in the code chunkS below.

```{r}
# creating a numerical vectors
vec = c(1,2,3,4)
vec
```

`Hint: To run a code chunk, click on the green play button at the top right of the code chunk.`

```{r}
# creating a vector with character data types
vec2 = c("R","Python","C++","JavaScript")
vec2
```


```{r}
# creating a vector with logical data types
vec3 = c(TRUE,FALSE,FALSE,TRUE)
vec3
```

### Using the seq() function.

```{r}
# creating numerical vector with seq
vec4 = seq(0,5)
vec4

# creating numerical vector with seq with a difference of 2 between each values
vec5 = seq(0,5, by=2)
vec5

# creating a vector with a defined length
# trying changing the length.out and observe
vec6 = seq(0,5, length.out=10)
vec6
```

### Using the colon operator

```{r}
# creating a numerical vector with values from 1 to 10
vec7 = 1:10
vec7

# creating a numerical vector with values from -1 to 10
vec8 = -1:10
vec8
```


`TIP: WHEN YOU LOOK oN YOUR ENVIRONMENT TAB, YOU SHOULS BE ABLE TO SEE ALL THE VECTORS WE HAVE CREATED`

## Vector Operations.

There are various operations that can applied to vectors, they include **vector length**, **vector arithmetics**, **combining vector** and **vector indexing**

### Vector Length

To find out the number of items in a vector, we use the `length()` function. For instance, we can use this function to find out how many items in the vectors we have created above as shown below, run the code chunk below to get the length of the vectors.

```{r}

```


# Bonus

## Comments

To write comments in R, you will use `#` followed by the code. The code chunk below illustrates how to write code. 

```{r}
# this is a comment, it will not be executed.
```


