---
title: "Fundermentals of R"
subtitle: "Advanced R - 1.0 Data Cleaning"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Data cleaning is an essential step in the data analysis process. It involves identifying and handling various data issues, such as missing values, duplicates, incorrect data types, and multi-collinearity. In this lesson, we will cover techniques for cleaning and preparing your data in R.

# Missing Data

## Checking Missing Data

Missing data can occur when certain observations or variables have incomplete or unknown values. It is crucial to identify and handle missing data appropriately to avoid biased analysis. R provides functions to detect and handle missing values, such as `is.na()` and `complete.cases()`.

Code Example:

```{r missing-data}
# clean data
data <- data.frame(
  x = c(1, 2, 3, 4, 5),
  y = c("a", "b", "c", "d", "e")
)

# Checking for missing values
is.na(data)
```


```{r missing-data-2}
# data with NA values
data2 <- data.frame(
  a = c(100, 203, 340, 450, 550),
  x = c(1, 2, 3, 4, NA),
  y = c(80, 83, NA, 90, 34),
  z = c(10, 20, 30, 40, 50)
)

# Checking for missing values
is.na(data2)
```

## Handling Missing Data

Handling missing data or NA values is crucial in data analysis to ensure accurate and reliable results. In R, there are several ways to handle missing data. Here are some common techniques

### Removing Missing Data

One approach is to remove the rows or columns with missing data from the dataset. This method is suitable when the missing data is random and does not significantly impact the overall analysis.

```{r omit-na}
# Removing rows with missing values
clean_data = na.omit(data2) 

# Removing columns with missing values
clean_data2 <- data2[, -c(2,3)] 

is.na(clean_data)
is.na(clean_data2)
```

### Imputing Missing Data with mean/median

Imputation involves replacing missing values with estimated or plausible values. This approach helps retain the complete dataset while minimizing the impact of missing data.

```{r impute-na}
data2_copy = data2
# Imputing missing values with mean
data2_copy$x <- ifelse(is.na(data2_copy$x), 
                       mean(data2_copy$x, na.rm = TRUE), data2_copy$x)

data2_copy$y <- ifelse(is.na(data2_copy$y), 
                       mean(data2_copy$y, na.rm = TRUE), data2_copy$y)

is.na(data2_copy)
```

# Duplicate Data

Duplicate data refers to observations that have identical values across multiple variables. It is important to identify and remove duplicates to avoid skewed analysis results. R offers functions like `duplicated()` and `unique()` for checking and handling duplicate data.

## Checking Duplicate Data

```{r duplicate-data}
# Creating a sample data frame with duplicate values
data_dup <- data.frame(
  x = c(1, 2, 3, 4, 2),
  y = c("a", "b", "c", "d", "b"),
  z = c(1, 2, 3, 4, 2)
)

# Checking for duplicate rows
any_dup <- any(duplicated(data_dup))
any_dup
```

## Count Duplicate rows

Sometimes, it may be useful to identify and count the number of duplicate rows in a dataset without removing them. This can help gain insights into the frequency of duplicate observations.

```{r dup-count}
# Counting duplicate rows
dup_count <- table(duplicated(data_dup))
dup_count
```

## Remove Duplicate Data

```{r remove-dup}
# get unique data
unique_data <- unique(data_dup)

any(duplicated(unique_data))
```

These techniques help in handling duplicate data in R. It is important to carefully consider the nature of the data and the specific analysis requirements when deciding on the appropriate method for handling duplicates.

# Dropping Columns/Rows

Removing unnecessary columns or rows in R involves excluding specific variables or observations from a dataset. Some common techniques for removing unnecessary columns or rows include: -

## Removing Columns by Index

```{r rem-col-by-index}
# make a copy of duplicated data frame
data_dup_copy = data_dup

# Removing columns 1
new_data_dup_copy <- data_dup_copy[, -1]

new_data_dup_copy
```

## Removing Columns by Name

```{r rem-col-by-name}
# make a copy of duplicated data frame
data_dup_copy = data_dup

# Removing columns 1
new_data_dup_copy <- data_dup_copy[, c("y","z")]

new_data_dup_copy
```

We can simplify the above process by using a package called `dplyr` as shown below

```{r rem-col-dplyr, warning=FALSE}
library(dplyr) # provides the select function
library(magrittr) # for piping operator

# remove column x
new_data_dup_copy =  data_dup %>%
  select(-x)

new_data_dup_copy
```


