---
title: "Fundermentals of R"
subtitle: "Intermediate R - 4.0 Hypothesis Testing"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output:   
  html_document: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Hypothesis testing is a statistical technique used to make inferences about a population based on sample data. In R, there are several functions and tests available for hypothesis testing. In this lesson, we will cover the following hypothesis tests: one-sample t-test, two-sample t-test, normality test, confidence interval, sign test, Wilcoxon signed-rank test, and Kruskal-Wallis rank sum test.

## One-Sample t-test

The one-sample t-test is used to determine if the mean of a sample significantly differs from a specified value.
    
The p-value obtained from the one-sample t-test determines whether the mean of the sample significantly differs from the specified value (mu = 27). If the p-value is less than the significance level (commonly 0.05), we reject the null hypothesis (H0) and conclude that the mean is significantly different.


      H0: the mean of the data is 27
      H1: the mean is not equal to 27

```{r one-sample-ttest}
# Sample Data
data <- c(23, 25, 26, 24, 28, 27, 22, 26, 25, 24)

# One-sample t-test
t.test(data, mu = 27)
```

### Conclusion

Since the p-value is less than 0.05, we reject the null hypothesis, Therefore, we can conclude that the mean of the data is not equal to 27.

## Two-Sample t-test

The two-sample t-test is used to compare the means of two independent samples.
        
The p-value obtained from the two-sample t-test determines whether the means of the two groups significantly differ. If the p-value is less than the significance level (0.05), we reject the null hypothesis and conclude that there is a significant difference between the means of the two groups.

          H0: the mean difference is zero (mean1 = mean2)
          H1: the mean difference is not zero (mean1 != mean2)

```{r two-sample-ttest}
# Sample Data
data1 <- c(23, 25, 26, 24, 28)
data2 <- c(27, 22, 26, 25, 24)

# Two-sample t-test
t.test(data1, data2)
```


### Conclusion

Since the p-value is greater than 0.05, we fail to reject the null hypothesis since we do not have enough evidence.

## Normality Test

The normality test is used to assess whether a sample follows a normal distribution.

The Shapiro-Wilk test is commonly used to check for normality. The p-value obtained from the test indicates whether the sample deviates significantly from a normal distribution. If the p-value is less than the significance level, we reject the null hypothesis and conclude that the data is not normally distributed.

          H0: the data is normally distributed
          H1: the data is not normally distributed
          
```{r shapiro-test}
# Sample Data
data <- rnorm(100)

# Normality test
shapiro.test(data)
```

### Conclusion

Since the p-value is greater than 0.05, we fail to reject the null hypothesis since we do not have enough evidence.

## Wilcoxon Signed-Rank Test

The Wilcoxon signed-rank test is a non-parametric alternative to the paired t-test when the data is not normally distributed.

The p-value obtained from the Wilcoxon signed-rank test determines whether there is a significant difference between the paired observations. If the p-value is less than the significance level, we reject the null hypothesis and conclude that there is a significant difference.

      H0: The median of the data of differences between the paired data is zero.
      H1: The median of the data of differences between the paired data is not zero
      
      
```{r wilcox-test, warning=FALSE}
# Sample Data
x <- c(23, 25, 26, 24, 28)
y <- c(27, 22, 26, 25, 24)

# Wilcoxon signed-rank test
wilcox.test(x, y, paired = TRUE)
```

### Conclusion

Since the p-value is 1, we fail to reject the null hypothesis and conclude that there is no significant difference between the paired observations. The test did not find any evidence of a significant difference between the paired observations.

## Kruskal-Wallis Rank Sum Test

The Kruskal-Wallis rank sum test is a non-parametric test used to compare the distributions of two or more independent groups.

The p-value obtained from the Kruskal-Wallis test determines whether there is a significant difference between the distributions of the groups. If the p-value is less than the significance level, we reject the null hypothesis and conclude that there is a significant difference.

      H0: The mean ranks of the groups are the same
      H1: The mean ranks of the groups are different
      
```{r kruskal-wallis}
# Sample Data
x <- c(23, 25, 26, 24, 28)
y <- c(27, 22, 26, 25, 24)
z <- c(29, 31, 30, 28, 27)

# Kruskal-Wallis test
kruskal.test(list(x, y, z))
```

### Conclusion

Since the p-value is less than 0.05, we reject the null hypothesis and conclude that the mean ranks of the groups are different.

HURRAY! YOU HAVE FINISHED THE LESSON, PROCEED TO THE ASSIGNMENT.

