---
title: "Fundermentals of R"
subtitle: "Intermediate R - 3.0 Probability Distributions"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Probability distributions are a fundamental concept in statistics and data analysis. In R, there are various built-in functions and packages that allow for the generation and manipulation of probability distributions. In this lesson, we will explore some commonly used probability distributions in R and demonstrate how to work with them in R.

# Probability Distributions

## Discrete Probability Distributions

### Binomial Distribution

The binomial distribution models the number of successes in a fixed number of independent Bernoulli trials. In R, the `rbinom()` function generates random numbers from a binomial distribution.

```{r r-binom}
# Generating random numbers from a binomial distribution
n <- 10  # number of trials

p <- 0.5  # probability of success on each trial

binom <- rbinom(n, size = n, prob = p)

binom
```

###  Poisson Distribution

The Poisson distribution models the number of events occurring in a fixed interval of time or space. In R, the `rpois()` function generates random numbers from a Poisson distribution.

```{r r-pois}
# Generating random numbers from a Poisson distribution
lambda <- 2  # average rate of events

poisson <- rpois(n, lambda = lambda)

poisson
```

## Continuous Probability Distributions

### Normal Distribution

The normal distribution, also known as the Gaussian distribution, is widely used in statistics and data analysis. In R, the `rnorm()` function generates random numbers from a normal distribution.

```{r r-norm}
# Generating random numbers from a normal distribution
mean <- 0  # mean of the distribution

sd <- 1  # standard deviation of the distribution

normal <- rnorm(n, mean = mean, sd = sd)

normal
```

### Uniform Distribution

The uniform distribution represents a continuous probability distribution where all outcomes are equally likely. In R, the `runif()` function generates random numbers from a uniform distribution.

```{r r-unif}
# Generating random numbers from a uniform distribution
min <- 0  # minimum value

max <- 1  # maximum value

uniform <- runif(n, min = min, max = max)

uniform
```

## Additional Distributions

### Exponential Distribution

```{r r-exp}
# Generating random numbers from an exponential distribution
rate <- 0.5  # rate parameter

exponential <- rexp(n, rate = rate)

exponential
```

### Gamma Distribution

```{r r-gamma}
# Generating random numbers from a gamma distribution
rate <- 0.5  # rate parameter

shape <- 0.5  

gamma <- rgamma(n, shape = shape,rate = rate)

gamma
```

### Chi-Square Distribution

```{r r-chisq}
# Generating random numbers from a chi-square distribution
df <- 5  # degrees of freedom

chi_square <- rchisq(n, df = df)

chi_square
```

HURRAY! YOU HAVE FINISHED THE LESSON, PROCEED TO THE ASSIGNMENT.
