---
title: "Fundermentals of R"
subtitle: "Intermediate R - 3.0 Probability Distributions Assignment"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Generate random numbers from a uniform distribution between -1 and 1 in R, and calculate the mean and standard deviation of the generated values.

```{r q1}
# your code
```

## Question 2

Plot the probability mass function (PMF) of a binomial distribution with n = 50 and p = 0.4 using R's base graphics system.

```{r q2}
# your code
```

## Question 3

Calculate the quatiles of a Poisson distribution with λ = 2.5 in R.

```{r q2}
# your code
```

## Question 4

Generate random numbers from a binomial distribution with parameters n = 50 and p = 0.3 in R and print the numbers.

```{r q4}
# your code
```

Use the vector you created in question 4 above, answer the questions below:

### Question 5

Generate the summary and write a brief history.

```{r q5}
# your code
```

### Question 6

Calculate the average and median of the vector. 

```{r q6}
# your code
```


### Question 7

Plot the density plot of the vector using the base functions.

```{r q7}
# your code
```

### Question 8

Calculate the minimum and maximum values of the vector.

```{r q8}
# your code
```

### Question 9

print out the 10th, 20th, 60th and 90th quantiles of the vector.

```{r q9}
# your code
```

## Question 10

Calculate the value of the CDF at x = 3.

```{r}
# your code
```

HURRAY! YOU HAVE FINISHED THE ASSIGNMENT, PROCEED TO THE NEXT LESSON.
