---
title: "Fundermentals of R"
subtitle: '4.0 Functions in R Assignment'
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up Question 1

Create a function that will accept the first and last name with the age of a person.

```{r}
# your function goes here
```


## Question 1

Print out the first name, last name and age of the person.

```{r}
# your code
```

## Question 2

Recreate the function and supply default values for the first name, last name and the age. Print the default values to the console.

```{r}
# your code
```

# Set Up Question 2

Create a function, *sample_num*, that will sample 20 numbers from 100 numbers. The function will first generate the 20 numbers and later on we will later on create a function to calculate the mean of the numbers. *(DO NOT USE THE BUILD IN MEAN FUNCTION)*

```{r}
# your code
```


## Question 1

Print out the numbers you generated in the set up above.

```{r}
# your code
```

## Question 2

Create a function, *sample_mean*, that accepts a list of numbers and calculates the mean of those numbers. The function should return the mean of the numbers.

*HINT:TEST THE FUNCTION ON SMALL NUMBERS FIRST*

```{r}
# your function goes here
```

## Question 3

Calculate the mean of the numbers you generated earlier using the function above.

```{r}
# your code
```

# Set Up Question 3

Write a function, *sum_n*, that will accept the number, n, which is the maximum number you will want to sum up-to. The function should return the sum of those numbers.

```{r}
# your code
```

## Question 1

What is the sum of the first 20 numbers?

```{r}
# your code
```

## Question 2

What is the sum of the first 100 numbers?

```{r}
# your code
```

## Question 3

What is the sum of the first 1050 numbers?

```{r}
# your code
```

# Conclusion

This is the end of the assignment, Proceed to the next lesson.

GOOD LUCK!!