---
title: "Fundermentals of R"
subtitle: 'Intermediate R - 1.0 Data'
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

R provides a wide range of built-in datasets that are readily available for analysis and exploration. These datasets cover various domains, including statistics, economics, biology, and social sciences. In this lesson, we will explore some commonly used built-in datasets in R along with code examples to access and manipulate them

## In Built Data

R comes with a package called "datasets" that contains numerous built-in datasets. To access these datasets, we have to load the "datasets" package. To view a list of all the datasets, you can use either `data()` or `library(help = "datasets")`, run the code chunks below to see them in action.

NOTE: This lesson is only limited to showing how to loading data, manipulation will come later on.

```{r method1}
library(help = "datasets")
```


```{r method2}
data()
```

### Sample npk Data

```{r npk-data}
# view the first 4 rows
head(npk, 4)
```

### Sample Cars Data

```{r cars-data}
# view the first 4 rows
head(cars, 4)
```