---
title: "Fundermentals of R"
subtitle: "Intermediate R - 1.0 Data Assignment"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

In this assignment, you will create all the datasets and then perform the required task.

## Question 1

Load the "iris" dataset from the built-in datasets in R and assign it to a variable called "iris_data". Display the first 5 rows of the dataset.

```{r q1}
# your code
```

## Question 2

You have an Excel file named `sales.xlsx` with multiple sheets. Load the sheet named `SalesData` into R and assign it to a variable called `sales_data`.

```{r q2}
# your code
```

## Question 3

Connect to a SQLite database named `mydatabase.db` and execute a query to retrieve all records from the `customers` table. Assign the result to a variable called `customer_data`.

```{r}
# your code
```

## Question 4

Load a text file named `notes.txt` and count the total number of lines in the file.

```{r q4}
# your code
```

## Question 5

Create a 3x3 matrix named `mat1` with the following values:

      1 2 3
      4 5 6
      7 8 9
      
Perform matrix multiplication of `mat1` with its transpose and assign the result to a new matrix called `mat2`. Display the contents of `mat2`.

```{r q5}
# your code
```

## Question 6

Create a matrix named `mat2` with the following values:

      2  4  6
      8  10 12
      14 16 18
      
Calculate the sum of the elements in each column of `mat2` and store the result in a vector called `column_sums`. Display the contents of `column_sums`.

```{r q6}
# your code
```

## Question 7

Create a 10*10 multiplication table named `Mult` using matrix and a for loop. Display the contents of `Mult`.  

```{r q7}
# your code
```

## Question 8

Create a data frame named `students` with the following columns: `Name`, `Age`, `Gender`, and `Grade`, Add at least five rows of data to the data frame.

```{r q8}
# your code
```

### Question 9

Extract the `Age` column from the `students` data frame and assign it to a variable called `ages`. Calculate the average age of the students.

```{r q9}
# your code
```

### Question 10

Filter the `students` data frame to include only female students. Assign the filtered data frame to a variable called `female_students`.

```{r q10}
# your code
```

### Question 11

Add a new column named `Pass` to the `students` data frame. The `Pass` column should indicate whether the student passed or failed based on a minimum passing grade. Display the updated data frame.

```{r q11}
# your code
```

### Question 12

Rename the `Grade` column in the `students` data frame to `FinalGrade`. Display the updated data frame structure.

```{r q12}
# your code
```

NICE ONE! YOU HAVE COMPLETED THE ASSIGNMENT, PROCEED TO THE STATISTICAL ANALYSIS LESSON.
