---
title: "Fundermentals of R"
subtitle: "Advanced R - 4.0 Machine Learning - Decision Tree"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# loading libraries
library(tidyverse)
library(mice) # missing data imputation
library(caret) # data splitting
library(party)
```

## Introduction

Decision Trees are powerful machine learning algorithms that can handle both regression and classification tasks. In this lesson, we will explore how to implement machine learning using Decision Trees in R, using the Titanic dataset as an example. We will cover the entire process, including data loading, cleaning, visualization, splitting, model training, evaluation, and making predictions.

## The Data

The sinking of the Titanic is one of the most infamous shipwrecks in history.

On April 15, 1912, during her maiden voyage, the widely considered "unsinkable" RMS Titanic sank after colliding with an iceberg. Unfortunately, there weren't enough lifeboats for everyone onboard, resulting in the death of 1502 out of 2224 passengers and crew.

While there was some element of luck involved in surviving, it seems some groups of people were more likely to survive than others.

In this challenge, we will build a predictive model that answers the question: "what sorts of people were more likely to survive?" using passenger data (ie name, age, gender, socio-economic class, etc). We will use the titanic dataset in this lesson, you can learn more from the kaggle website.

## Data Loading

The dataset consists of the train and test datasets. We will use the training dataset to train and test our model and the best model will be used to make final predictions using the test set.

The data dictionary:

| Variable | Definition                                | Key                                            |
|---------------|---------------------------|------------------------------|
| Survival | Survival                                  | 0 = No, 1 = Yes                                |
| pclass   | Ticket class                              | 1 = 1st, 2 = 2nd, 3 = 3rd                      |
| Sex      | Sex                                       | male, female                                   |
| Age      | Age in years                              |                                                |
| sibsp    | \# of siblings/spouses aboard the Titanic |                                                |
| parch    | \# of parents/children aboard the Titanic |                                                |
| ticket   | Ticket number                             |                                                |
| fare     | Passenger fare                            |                                                |
| cabin    | Cabin number                              |                                                |
| embarked | Port of Embarkation                       | C = Cherbourg, Q = Queenstown, S = Southampton |

: Data Dictionary

```{r data-loading}
# loading the train data
train = read.csv("data/train.csv")

# subset columns you need
train_updated = train %>%
  select(-c(PassengerId, Name, SibSp, Parch, Ticket, Cabin, Embarked))

# data summary
summary(train_updated)

# na count
colSums(is.na(train_updated))
```

## Data Cleaning

### Missing Data

```{r handle-na}
# define the imputation model
imp_model <- mice(train_updated, method = "pmm", m = 5)

# get complete data
imputed_data <- complete(imp_model)

colSums(is.na(imputed_data))
```

### Data Types

The data types of all variables are wrongly labeled except for Age and Fare. We will convert the data types of Survived, PClass and Sex into factors.

```{r data-types}
# converting data types to factors
imputed_data = imputed_data %>%
  mutate(
    Survived = as.factor(Survived),
    Pclass = as.factor(Pclass),
    Sex = as.factor(Sex)
  )


# data summary
summary(imputed_data)
```

## Exploratory Data Analysis

Since we have done EDA using this data in both logistic and random forest lessons, we will not repeat but you can plot different plots from the ones we have done.

```{r EDA}
# your plots
```

## Modelling

### Splitting the Data

We will split the data into: -

      1. Training Data (80%).
      2. Validation Data (20%).
      
```{r data-splitting}
# splitting index
index = createDataPartition(imputed_data$Survived, p=0.8, list = FALSE)

# training data
training = imputed_data[index,]

# validation data
validation = imputed_data[-index,]
```

### Fitting Base model

```{r base-model}
# Decision Tree model training
model <- ctree(Survived~., data = training)

# model plot
plot(model)
```

### Predictions and Evaluation

```{r pred}
# predictions
pred = predict(model, newdata = validation, type = "response")

# confusion matrix
mt = table(validation$Survived,pred)

confusionMatrix(mt)
```

The base model achieved an accuracy of `r confusionMatrix(mt)$overall[['Accuracy']]*100`.

CONGRATULATIONS ON FINISHING YOUR LESSON, PLEASE PROCEED TO THE ASSIGNMENT.