---
title: 'Fundermentals of R'
subtitle: '2.0 Data Structures Assignement'
author: "Prepared By Clinton Moshe Mwachia"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions.

      1. Comment on what your code does.
      2. Make sure the code only prints the answer.

# Vectors

## PREPARATION.

Create two vectors, **animals**, containing the big five animals in KENYA and **marks** containing the math scores of five candidates. Use the code chunk below. 

```{r prep}

```

## Question 1

What is the length of the vector **animals**?

write your code below.

```{r}

```

## Question 2

What is the length of the vector **marks**?

write your code below.

```{r}

```

## Question 3

Extract the 1st, 3rd and 5th animals and print them to the console.

```{r}

```

## Question 4

What is the highest math score of the candidates?

*hint: sort the marks in descending order and extract the first item*

```{r}

```

# Lists

## Preparation.

create a list, **my_list** that contains the two vectors you created earlier.

```{r}

```

## Question 1

What is the length of the list you have created above?

```{r}

```

## Question 2

create a new vector, **vec** containing numbers 100 and 200 and append it to **my_list**.

What is the length of the new list?

```{r}

```

## Question 3

The new list created above does not have names.Proceed to the code chunk below and appropriately name the items. Print the new list.

```{r}

```

## Question 4

Extract the 1st, 3rd and 5th animals from the list above and print them to the console

```{r}

```

# Matrices

## Preparation

create a 2 by 3 matrix **A** containing the data, `1,2,3,4,5,6`, filed by row. Print the matrix

```{r}

```

## Question 1

What are the dimensions of matrix A?

```{r}

```

## Question 2

Matrix A does not have column and row names, using appropriate functions, confirm this and name both the rows and columns of the matrix. Print the matrix

```{r}

```

## Question 3

What is A + A? Print the matrix

```{r}

```

## Question 4

What is A - A? Print the matrix

```{r}

```

## Question 5

What is A + 2A? Print the matrix

```{r}

```

## Question 6

What is the transpose of matrix A?

```{r}

```

# Data Frames

## Preparation

Create two data frames, **df & df2**. Data frame df should contain column named *one* with the data `10.5,11,23,4` and df2 with column named two containing data `1,3,5,7`.

Use the code chunk below.

```{r}
df = data.frame(one=c(10.5, 11, 23, 4))

df

df2 = data.frame(two=c(1, 3, 5, 7))

df2
```

## Question 1

Print the structures of the data frames.

```{r}
str(df)
```

## Question 2

What is the min, mean, max, and 3rd quartile of column one and two?

*hint: get the summary of the data frames*

```{r}

```

## Question 3

Rename both column `one -> One` and column `two -> Two`.

Print out the data frames.

```{r}

```

## Question 4

Add a new column to the data frame df and fill it with the data from data frame df2. The new column name be named Two.

Print out the new data frame df.

```{r}

```

## Question 5

Combine the two data frames using the appropriate function and assign it to **new_df**

What is new dimensions of new_df?

```{r}

```

# THE END!!