---
title: "Fundermentals of R"
subtitle: "Intermediate R - 5.0 Data Visualization"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Data visualization plays a crucial role in understanding and interpreting data. In R, there are various graphical techniques available to visualize data effectively. In this lesson, we will explore six commonly used visualization methods: scatter plot, line plot, pie chart, bar plot, box-plot, and histogram. 

## The data

we will create a data frame that we will use for all the visuals

```{r data}
# the data
data = data.frame(
  subject = c("Math", "Eng", "Kis", "Chem", "Phy", "Hist", "Geo", "Bio", "CRE", "IRE"),
  marks = c(45, 50, 52, 59, 55, 59, 60, 62, 79, 85),
  age = c(15, 18, 25, 45, 49, 45, 58, 65, 69, 79),
  gender = c("M","F","M","F","F","F","M","M","F","M")
)

data
```

## Scatter Plot

A scatter plot is used to visualize the relationship between two continuous variables. It displays individual data points as dots on a two-dimensional plane.

```{r scatter-plot}
# Scatter plot
plot(data$age, data$marks, main = "Age vs Marks", xlab = "Age", ylab = "Marks")
```

### Interpretation

The scatter plot shows a positive linear relationship between the Age and Marks. As Age increases, Marks also increases, indicating a strong positive correlation.

## Line Plot

A line plot is commonly used to visualize the trend or pattern of a variable over time or another continuous variable.

```{r line-plot}
# Line plot
plot(data$age, data$marks, type = "o", main = "Age vs Marks", 
     xlab = "Age", ylab = "Marks")
```

### Interpretation

The line plot shows the trend of the Marks over Ages. It indicates an initial increase, followed by a slight decrease and then a further increase.

## Pie Chart

A pie chart is a circular visualization that represents categorical data as slices of a pie, with each slice representing a proportion of the whole.

```{r pie-chart}
# pie chart
pie(data$marks, labels = data$subject, main = "Marks")
```

### Interpretation

The pie chart displays the marks data as proportions of the whole. Each slice represents a subject, and the size of the slice corresponds to the proportion of that subject.

## Bar Plot

A bar plot is used to compare different categories or groups by displaying bars of varying lengths.

```{r bar-plot}
# Bar plot
barplot(data$marks, names.arg = data$subject, main = "Subject Performance", 
        xlab = "Subject", ylab = "Marks")
```

### Interpretation

The bar plot visually compares the marks of different subject. IRE has the highest marks and Math has the lowest.

## Boxplot 

A boxplot summarizes the distribution of a continuous variable by displaying the median, quartiles, and outliers.

```{r box-plot}
# boxplot
boxplot(data$marks~data$gender, main = "Distribution of Marks by gender", ylab = "Marks")
```

### Interpretation

The boxplot shows the distribution of marks in each gender. It displays the median, quartiles, and any outliers. Male students have the highest mean value, followed by female.

## Histogram

A histogram visualizes the distribution of a continuous variable by grouping the data into bins and displaying the frequencies or densities of each bin.

```{r hist}
# Histogram
hist(data$marks, main = "Marks", xlab = "Marks", ylab = "Frequency")
```

### Interpretation

The histogram displays the frequency of values in different bins. It provides insights into the distribution and shape of the marks data. In this example, the data appears to be somewhat right-skewed and most of the student got between 50 and 60.

HURRAY! YOUR DONE WITH THE LESSON, PLEASE PROCEED TO THE ASSIGNMENT.
