---
title: "Fundermentals of R"
subtitle: '4.0 Functions in R'
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

A function in R is a block of code that performs certain tasks and can only run when it is called. Functions are normally created and stored into libraries. In this session, we will learn about the following: -

      1. Built on functions.
      2. User defined functions.
      
# Built in functions

These are functions that can be found in the default R work-space. They come with the R and RStudio ecosystem and can be accessed without installing any libraries. Some of the built functions in R are `mean`, `mode`,`median`, `sum` and many more, The code chunks below will show some of the built in functions.

## Absolute value of a number

```{r}
# get the absolute value of a number
abs(-10)
```

## sum of 10 numbers

```{r}
# initialize the sum
total = 0 
# loop through 10 numbers and sum them
for (i in 1:10) {
  total = total + i
}

# print the total
cat(total)
```

## Logarithm of a number

```{r}
log(2)
```

## Split a string by space

```{r}
# define the string
name = "James clinton"

# split the string by space
strsplit(name, " ")
```
