---
title: "Fundermentals of R"
subtitle: '4.0 Functions in R'
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

A function in R is a block of code that performs certain tasks and can only run when it is called. Functions are normally created and stored into libraries. In this session, we will learn about the following: -

      1. Built on functions.
      2. User defined functions.
      
# Built in functions

These are functions that can be found in the default R work-space. They come with the R and RStudio ecosystem and can be accessed without installing any libraries. Some of the built functions in R are `mean`, `mode`,`median`, `sum` and many more, The code chunks below will show some of the built in functions.

## Absolute value of a number

```{r}
# get the absolute value of a number
abs(-10)
```

## sum of 10 numbers

```{r}
# initialize the sum
total = 0 
# loop through 10 numbers and sum them
for (i in 1:10) {
  total = total + i
}

# print the total
cat(total)
```

## Logarithm of a number

```{r}
log(2)
```

## Split a string by space

```{r}
# define the string
name = "James clinton"

# split the string by space
strsplit(name, " ")
```

## The mean of numbers

```{r}
# the mean of 2 numbers
mean(c(1,3))
```

# User Defined Functions

These a functions created by the user to perform specific tasks defined by the user. They are mainly created so that they can be re-used. These kinds of functions can contain arguments, or they can be without arguments or they can have default arguments. Functions with arguments are defined using the format,

$$
      functionName = function (arguments) {
        return(res)
      }
$$
And to use the function created above, you just call it `functionName(parameters)`. The code chunk below will show how to create a function that will allows the user enter the name and the it will greet them.

## Function with arguments

```{r}
# defining the function
Hello <- function(name){
  cat(paste0("Hello, ", name))
}

# test case 1
Hello("clinton")

# test case 2
Hello("James Peter")
```

## Function without arguments

```{r}
# defining the function
Hello2 <- function(){
  cat("Hello, Clinton")
}

# test case 1
Hello2()
```

## Function with default arguments

```{r}
# defining the function
Hello3 <- function(name = "Clinton Moshe"){
  cat(paste0("\nHello, ", name))
}

# test case 1
Hello3()

# test case 2
Hello3("John Doe")
```

**NOTE: IF YOU WANT THE FUNCTION TO RETURN A VALUE,USE THE RETURN STATEMENT AS ILLUSTRATED BELOW**

The function will return the sum of 20 numbers.

```{r}
# defining the number
sum_20 <- function(){
  # initialize the total
  tot = 0
  # loop through 20 numbers and sum
  for (i in 1:20) {
    tot = tot + i
  }
  # return the value
  return(tot)
}

# calling the function
sum_20()
```

# Assignement

This is the end of the lesson, Proceed to the assignment.

GOOD LUCK!!